Fast and scalable Gaussian process modeling of stellar variability \vspace{-3\baselineskip}

### 2016-12-18 paper/ms.tex
Described SHO; added placeholder for Sturm's theorem.

%
% New Outline:
% 1. Introduction
% 2. Gaussian processes
%   - GPs in general
%   - RP method
%   - Ambikasaran method
%   - Generalization
% 3. Implementation and performance
%   - Implementation considerations: band solver vs. sparse, parameterization
%   - Benchmark numbers
%   - Scaling with N and M
%   - Description of APIs
% 4. Examples
%   - Simulated data: a simple exact GenRP simulation, a GenRP simulation with
%     structure, a few standard kernels, a AGW kernel
%   - An AGN light curve
%   - An RR Lyrae
%   - A kepler light curve
%   - An asteroseismic target
% 5. Comparisons to other methods
% 6. Summary
%   - Possible applications
%

@@

@@

>>>>

%
% New Outline:
% 1. Introduction
% 2. Gaussian processes
%   - GPs in general
%   - RP method
%   - Ambikasaran method
%   - Generalization
%   - Relation to stochastically-driven, damped simple harmonic oscillator
% 3. Implementation and performance
%   - Implementation considerations: band solver vs. sparse, parameterization
%   - Ensuring positive-definite covariance matrix: Sturm's theorem
%   - Benchmark numbers
%   - Scaling with N and M
%   - Description of APIs
% 4. Examples
%   - Simulated data: a simple exact GenRP simulation, a GenRP simulation with
%     structure, a few standard kernels, a AGW kernel
%   - An AGN light curve
%   - An RR Lyrae
%   - A kepler light curve
%   - An asteroseismic target
% 5. Comparisons to other methods
% 6. Summary
%   - Possible applications
%

@@

The general ACF is given by
\begin{equation}
K(\tau) = a_r e^{-b_r \vert\tau\vert} \cos{(b_r \vert\tau\vert)} + a_i e^{-b_r \vert\tau\vert} \sin{(b_r \vert\tau\vert)}.
\end{equation}
which has a power spectrum given by
\begin{equation}
P(\omega) = P_0 \frac{(a_ib_i+a_rb_r)(b_i^2+b_r^2)+(a_rb_r-a_ib_i)\omega^2}{(\omega^2-(b_i^2+b_r^2))^2 + 4b_r^2\omega^2}.
\end{equation}
The power spectrum should be non-negative since it is the absolute square of the Fourier transform of the noise.
Note that the denominator is always non-negative, so the power spectrum is non-negative if $a_ib_i+a_rb_r \ge 0$ and
$a_rb_r-a_ib_i \ge 0$.

\subsection{Relation to stochastically-driven, damped simple harmonic oscillator}

Intrinsic stellar variability consists primarily of random fluctuations driven by 
convection, oscillations excited by convection, and rotational variability.  These first
two types of variability may be well-described by a model of a damped simple-harmonic oscillator
driven by stochastic variability \citep{1990ApJ...364..699A}.  The governing differential equation is
\begin{equation}
y^{\prime\prime}+ \frac{\omega_0}{Q} y^\prime + \omega_0^2 y = \epsilon,
\end{equation}
where $y(t)$ is a displacement, $^{\prime}$ represents a time derivative,
$\omega_0$ is the frequency of an undamped oscillator, $Q$ is the quality factor
of the oscillation, and $\epsilon(t)$ is a stochastic driving force.

In the limit of an infinite time series and white random forcing, the noise-free power 
spectrum of this equation is given by:
\begin{equation}
P(\omega) = P_0 \frac{\omega_0^4}{(\omega^2-\omega_0^2)^2 + \omega_0^2\omega^2/Q^2},
\end{equation}
where $P_0$ is a normalization constant.

This power spectrum matches that of our basis kernel if $\omega_0^2 = b_r^2+b_i^2$,
$a_rb_r-a_ib_i = 0$, $a_ib_i+a_rb_r=\omega_0^2$, and $4b_r^2 = \omega_0^2/Q^2$.
The power spectrum can be normalized by $P_0$ in amplitude and $\omega_0$ in
frequency to yield a spectral shape that only depends upon $Q$.  This is plotted
in Figure \ref{sho_psd} for various values of $Q$.  This power spectrum has
a peak frequency given by $\omega_{max} = \omega_0\sqrt{1-1/(2Q^2)}$, which is
greater than zero for $Q > 1/\sqrt{2}$.

\begin{figure}[!htbp]
\begin{center}
\includegraphics[width=\textwidth]{./approximate_kernels/sho.png}
\caption{Normalized power spectra of the stochastically-driven, damped simple
harmonic oscillator.}
\label{sho_psd}
\end{center}
\end{figure}

This power spectrum has several limits of interest:
\begin{itemize}
\item $Q=1/2$:  This results in a power spectrum $\propto 1/(x^2+1)^2$, where $x=\omega/\omega_0$.
This matches the power spectrum of the Mat\'ern 3/2 kernel, and can be represented by the following
autocovariance function:
\begin{equation}
K(\tau) = a_r e^{-b_r \vert\tau\vert} \left(\cos{(b_i \vert\tau\vert)} + \frac{b_r}{b_i}\sin{(b_i \vert\tau\vert)}\right),
\end{equation}
in the limit that $b_i \rightarrow \infty$, for which the second term has a coefficients that diverges,
but with large values of $b_i$, this may approximate the Mat\'ern kernel well.  Note that this
power spectrum peaks at $\omega=0$.
\item $Q=1/\sqrt{2}$:  This results in a power spectrum $\propto 1/(x^4+1)$, which is a functional
form used to fit the Solar granulation power spectrum by \cite{2009A&A...495..979M}.  This kernel has
dependence
\begin{equation}
K(\tau) = a_r e^{-b_r\vert\tau\vert} \left(\cos{(b_i\vert\tau\vert)}+\sin{(b_i\vert\tau\vert)}\right).
\end{equation}
Despite the sinusoidal dependence of the kernel, the strong damping causes the power spectrum to
peak at zero frequency; such a low quality oscillator doesn't display even quasi-periodic behavior.
\item $Q \gg 1$: This represents a high-quality oscillator which approaches a narrow peak.  The
power spectrum approaches
\begin{equation}
P(\omega) \approx P_0 \frac{Q^2}{1-1/(4Q^2)+4x_{max}^2\left(x-x_{max}\right)^2Q^2},
\end{equation}
where $x_{max}=\sqrt{1-1/(2Q^2)}$.  This shape approaches a Lorentzian for $x \approx x_{max}$ with
a width of $1/(2Qx_{max})$ and a peak height $\propto Q^2$.
\end{itemize}

The stochastically driven, damped simple-harmonic oscillator can then represent a wide range of intrinsic stellar variability, with
low $Q \approx 1$ representing granulation noise, and high $Q >> 1$ representing asteroseismic variability.

@@

\subsection{Ensuring positive definiteness: Sturm's theorem}

Relation of positive power spectrum to positive definite covariance matrix.
Power spectrum of a sum of kernels.
Sturm's theorem applied to numerator of power spectrum.
